Dynamics of Laterally Propagating Flames in X-ray Bursts

### 2019-11-25 xrb/paper.tex
Some additional comments, edits for XRB paper

The first vertically resolved simulations of lateral deflagrations
were obtained by \citet{cavecchi:2013} who showed how the Coriolis
force creates a geometrical configurations that increases the flame
speed set by conduction by a factor $\sim ~ L_r / H$, where $L_r$ is the
Rossby radius and $H$ is the scale height of the burning layer. The
effect on flame propagation of changing Coriolis confinement across
the surface was explored in \citet{art-2015-cavecchi-etal}, while \citet{art-2016-cavecchi-etal}
showed how the magnetic field tension, opposing the Coriolis
force, can either speed up or slow down the flame by changing the
flame front orizontal extent. Finally, \citet{Cavecchi2019}
studied the effects in 3D of the baroclinic instability at the flame
front, measuring flames up to 10 times faster than in the 2D case.

@@

All simulations are performed with the \castro\ hydrodynamics
code~(\citealt{castro}; see also \citealt{astronum:2017} for a recent
description).  We evolve the system of fully compressible Euler
equations for reacting flow:
\begin{eqnarray}
\frac{\partial( \rho X_k)}{\partial t} &=& - \nabla\cdot (\rho \Ub X_k) + \rho \omegadot_k \\
%
\frac{\partial (\rho \Ub)}{\partial t} &=& - \nabla\cdot (\rho \Ub \Ub) - \nabla p +
    \rho \gb \nonumber \\
  &&-2 \rho \Omegab\times \Ub - \rho \Omegab \times (\Omegab \times \rb) \\
%
\frac{\partial (\rho E)}{\partial t} &=& - \nabla \cdot (\rho \Ub E + p\Ub) +
    \nabla \cdot \kth \nabla T + \rho \epsdot +\nonumber\\
  && \rho \Ub \cdot \gb - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho |\Omegab|^2 (\Ub \cdot \rb)
\end{eqnarray}
Here, $\rho$ is the mass density, $\Ub$ is the velocity, $p$ is the pressure and $E$ is
the specific total energy, which is related to the specific internal
energy as $e = E - |\Ub|^2/2$.  Species are described by mass
fractions, $X_k$ (such that $\sum_k X_k = 1$), and creation rates,
$\omegadot$, and are related to the total specific energy generation
rate, $\epsdot$.  The total mass conservation,
\begin{equation}
\frac{\partial \rho}{\partial t} = - \nabla\cdot (\rho \Ub)
\end{equation}
implies $\sum_k \omegadot_k = 0$.  \castro\ uses an unsplit piecewise
parabolic method (PPM) with characteristic tracing for solving the
hydrodynamics~\citep{ppm,millercolella:2002}, generalized to an
arbitrary equation of state~\citep{zingalekatz}.  Reactions are
incorporated via Strang splitting~\citep{strang:1968}, giving a method
that is overall second-order accurate in space and time.
\castro\ uses the \amrex\ adaptive mesh refinement
library~\citep{amrex_joss} to manage a hierarchy of grids at different
resolutions.

@@

Writing the momentum equation in terms of the $u$, $v$, and $w$
components, and neglecting the centrifugal force, we have:
\begin{align}
\frac{\partial (\rho u)}{\partial t} + \nabla \cdot (\rho u \Ub) +
     \frac{\partial p}{\partial x} &= -2\rho \Omega_0 w  \\
%
\frac{\partial (\rho v)}{\partial t} + \nabla \cdot (\rho v \Ub) +
     \frac{\partial p}{\partial y} &= -\rho g \\
%
\frac{\partial (\rho w)}{\partial t} + \nabla \cdot (\rho w \Ub) +
     \cancelto{0}{\frac{\partial p}{\partial z}} &=
    2\rho \Omega_0 u
\end{align}
where we cancel $\partial p/\partial z$ because there are no
variations in the azimuthal direction.  This allows us to recast the
$w$-velocity equation as a simple advection equation:
\begin{equation}
\frac{\partial w}{\partial t} + \Ub \cdot \nabla w = 2\Omega_0 u
\end{equation}
In our geometry, the flame will propagate from left to right, so $u$
will be positive and the Coriolis force results in $w > 0$ (out of the
simulation plane).  The algorithmic implementation of rotation in
\castro\ is described in \cite{wdmergerI}.

We use a general stellar equation of state with nuclei (treated as an
ideal gas), photons, and degenerate/relativistic electrons, as
described in \cite{timmes_swesty:2000}.  To model reactions, we use a 
13-isotope alpha chain, derived from the {\tt aprox13}
network~\citep{timmes_aprox13}.  For one of our runs, we use the
smaller 7-isotope network described in \citet{iso7}.  We integrate the
network using the VODE integration package~\citep{vode}, and our
implementation is provided in the StarKiller Microphysics
source~\citep{starkiller}.
We note that we do not explicitly model viscosity.  The reactions will
provide the smallscale cutoff to the instabilities and turbulence at
the flame front.  We also do not include species
diffusion---astrophysical flames tend to have large Lewis numbers, so
this is not expected to be important~\citep{timmeswoosley:1992}.
Finally, we use the thermal conductivities described in
\citet{Timmes00}.

@@

@@

We measure the flame width as:
\begin{equation}
\lambda_f \equiv \frac{\Delta T}{\max\{|\nabla T|\}}
\end{equation}
Experience with modeling resolved flames suggests that we need a
spatial resolution, $\Delta x$ of $\lambda_f/\Delta x \sim 5$~\citep{SNld}.  These
conditions represent the bottom of the He layer.  As the density
decreases with altitude, the flame thickness increases and the flame
speed decreases, so we will easily resolve the flame structure throughout
the rest of the atmosphere.

Flame propagation rates from both the 1D laminar flames and full 2D simulations are given in Table \ref{table:flame_speeds}.  \MarginPar{I think we should only report the laminar speeds here and defer the turbulent speeds until the section where we talk about how they are computed}

\begin{deluxetable}{lc}
	\tablecaption{\label{table:flame_speeds} Laminar flame speeds.}
	\tablehead{\colhead{run} & \colhead{$s_L$ (km s$^{-1}$)}}
	\startdata
	10/10 boost & $1.06 \pm 0.01$ \\
	5/5 boost & $0.56 \pm 0.01$ \\
	\enddata
	%
\end{deluxetable}

@@

Table~\ref{table:sim_names} summarizes the simulations that we ran. \MarginPar{This table has different model names to the previous table}
The majority of them used a reaction rate boosting of $10$ and a
conductivity boosting of $10$, which should increase the flame speed
by a factor of $10$.  Most simulations used a resolution of
$10~\mbox{cm}$ and a domain width of a little more than one kilometer.  
We use an artifically high rotation rate of
$2000~\mbox{Hz}$, which gives a Rossby length of
\begin{equation}
L_R = \frac{\sqrt{g H_0}}{\Omega} \sim 3\times 10^4~\mathrm{cm}
\end{equation}
using a scale height $H_0 = 10^3~\mathrm{cm}$.  This is about one
quarter of the domain width.  The run at $20~\mbox{cm}$ resolution used
one fewer level of refinement.  The slower rotating case using a
slightly wider domain to accommodate the expected larger Rossby
length.  We note that the entire simulation framework for these
calculations is freely available in the \castro\ github
repository\footnote{\url{https://github.com/AMReX-Astro/Castro}, using
  the {\tt flame\_wave} setup.}.  In the discussions below, we'll use
the simulation names defined in the table to refer to specific runs
and we'll use the 10/10 run as the reference calculation.

@@

Figure~\ref{fig:10_10_overview} shows the temperature, energy
generation rate, and $w$ component of the velocity (the out-of-plane
velocity induced by the Coriolis force).  This latter field
illustrates the hurricane effect set up by the laterally spreading
burning front.  In the energy generation rate plot, we see that the
burning is mostly concentrated toward the bottom of the layer, as
expected since the density is greatest there.  We see that the peak of
the burning has moved off of the symmetry axis, demonstrating that the
burning front is propagating to the right.  In the temperature plot,
we see the effect of our refinement criteria focusing only on the
atmosphere where we are most dense, with an artificial change in the
temperature at the refinement boundary due in part to the construction
of the initial model using the fine grid resolution for HSE. \MarginPar{Remind the reader what this
  was/where they should be looking in the plot to see this temperature
  drop} As we'll see when the lower resolution case, we do not believe
that this affects the results.

@@

We also considered the effect of the network size.
Figure~\ref{fig:network} shows a comparison of the 10/10 boosting run
with the standard 13-isotope ``aprox13'' network and the reduced
7-isotope ``iso7'' network.  We see that the flame in the aprox13 case
is slightly more advanced and has a higher $\bar{A}$ than the iso7
case.  We see in the next section that these two networks give largely
the same flame speed.  The reduced network size saves a lot of memory,
which will be useful when we transition to 3-d simulations.

@@

As we see in Figure~\ref{fig:flame_speed}, all of the various
approximations behave as we would expect.  The 1000~Hz run goes 2
times faster than the 2000~Hz run, as expected by the reduced rossby
radius \citep{cavecchi:2013}).  The 5/5 boost run is half the speed of
the 10/10 boost run, also as expected, consistent with the 1D laminar
flames and behaving as predicted by Eq.~\ref{eq:flame_scaling}.  This
confirms that the role of the Coriolis force is really to boost the
laminar flame by the geometrical/hydrodynamical effect expected ($v_f
= v_l * L_r / H$), similarly to what the 1000 Hz simulation
said.

@@

Baroclinicity is calculated as
\begin{equation}
    \boldsymbol{\psi} = \frac{1}{\rho^2} \nablab p \times \nablab \rho,
\end{equation}
and shows the misalignment of the local density and pressure gradients
\citep{Malone2014a}. As we are considering a 2d \MarginPar{We've used
  at least three different ways of writing 2-d/2D/2d so far. These
  should be changed so we used one style throughout.}system, the
component of the baroclinicity we consider here (out of the plane)
reflects the misalignment of the fields in the plane of the
simulation. Figure~\ref{fig:baroclinicity} shows that the
baroclinicity peaks along the flame front. This baroclinicity
generates vorticity, which in turn entrains material along the surface
of the flame (see also \citet{cavecchi:2013}). This same vorticity,
in 3D, should perturb the flame front and further increase flame speed
\citep{Cavecchi2019}.

@@

To test our implementation of diffusion, we created a unit test in
\castro\ that just uses the diffusion solver.  This test uses the
standard Gaussian initial conditions---the diffusion of a Gaussian
remains Gaussian, with a lower amplitude and wider width.  This is run
with a constant diffusion coefficent.  The point of this test is to
demonstrate that the implementation done in a predictor-corrector
form is second-order accurate.

>>>>

The first vertically resolved simulations of lateral deflagrations
were obtained by \citet{cavecchi:2013}, who showed how the Coriolis
force creates a geometrical configuration that increases the flame
speed set by conduction by a factor $\sim ~ L_R / H$, where $L_R$ is the
Rossby radius and $H$ is the scale height of the burning layer. The
effect of changing Coriolis confinement across the surface on the flame 
propagation was explored in \citet{art-2015-cavecchi-etal},
while \citet{art-2016-cavecchi-etal} showed how magnetic field tension,
opposing the Coriolis force, can either speed up or slow down the flame by
changing the horizontal extent of the flame front. Finally, \citet{Cavecchi2019}
studied the effects in 3D of the baroclinic instability at the flame
front, measuring flames up to 10 times faster than in the 2D case.

@@

All simulations are performed with the \castro\ hydrodynamics
code~(\citealt{castro}; see also \citealt{astronum:2017} for a recent
description).  We evolve the system of fully compressible Euler
equations for reacting flow:
\begin{eqnarray}
\frac{\partial( \rho X_k)}{\partial t} &=& - \nabla\cdot (\rho \Ub X_k) + \rho \omegadot_k \\
%
\frac{\partial (\rho \Ub)}{\partial t} &=& - \nabla\cdot (\rho \Ub \Ub) - \nabla p +
    \rho \gb \nonumber \\
  &&-2 \rho \Omegab\times \Ub - \rho \Omegab \times (\Omegab \times \rb) \\
%
\frac{\partial (\rho E)}{\partial t} &=& - \nabla \cdot (\rho \Ub E + p\Ub) +
    \nabla \cdot \kth \nabla T + \rho \epsdot +\nonumber\\
  && \rho \Ub \cdot \gb - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho |\Omegab|^2 (\Ub \cdot \rb)
\end{eqnarray}
Here, $\rho$ is the mass density, $\Ub$ is the velocity, $p$ is the pressure and $E$ is
the specific total energy, which is related to the specific internal
energy as $e = E - |\Ub|^2/2$.  Species are described by mass
fractions, $X_k$ (such that $\sum_k X_k = 1$), and creation rates,
$\omegadot$, and are related to the total specific energy generation
rate, $\epsdot$\MarginPar{Is notation consistent between E, e, epsilondot, and enucdot?}.  The total mass conservation,
\begin{equation}
\frac{\partial \rho}{\partial t} = - \nabla\cdot (\rho \Ub)
\end{equation}
implies $\sum_k \omegadot_k = 0$.  \castro\ uses an unsplit piecewise
parabolic method (PPM) with characteristic tracing for solving the
hydrodynamics~\citep{ppm,millercolella:2002}, generalized to an
arbitrary equation of state~\citep{zingalekatz}.  Reactions are
incorporated via Strang splitting~\citep{strang:1968}, giving a method
that is overall second-order accurate in space and time.
\castro\ uses the \amrex\ adaptive mesh refinement
library~\citep{amrex_joss} to manage a hierarchy of grids at different
resolutions.

@@

Writing the momentum equation in terms of the $u$, $v$, and $w$
components, and neglecting the centrifugal force, we have:
\begin{align}
\frac{\partial (\rho u)}{\partial t} + \nabla \cdot (\rho u \Ub) +
     \frac{\partial p}{\partial x} &= -2\rho \Omega_0 w  \\
%
\frac{\partial (\rho v)}{\partial t} + \nabla \cdot (\rho v \Ub) +
     \frac{\partial p}{\partial y} &= -\rho g \\
%
\frac{\partial (\rho w)}{\partial t} + \nabla \cdot (\rho w \Ub) +
     \cancelto{0}{\frac{\partial p}{\partial z}} &=
    2\rho \Omega_0 u
\end{align}
where we cancel $\partial p/\partial z$ because there are no
variations in the azimuthal direction.  This allows us to recast the
$w$-velocity equation as a simple advection equation:\MarginPar{In conjunction with total mass conservation?}
\begin{equation}
\frac{\partial w}{\partial t} + \Ub \cdot \nabla w = 2\Omega_0 u
\end{equation}
In our geometry, the flame will propagate from left to right, so $u$
will be positive and the Coriolis force results in $w > 0$ (out of the
simulation plane).  The algorithmic implementation of rotation in
\castro\ is described in \cite{wdmergerI}.

We use a general stellar equation of state with nuclei (treated as an
ideal gas), photons, and degenerate/relativistic electrons, as
described in \cite{timmes_swesty:2000}.  To model reactions, we use a 
13-isotope alpha chain network derived from the {\tt aprox13}
network~\citep{timmes_aprox13}.  For one of our runs, we use the
smaller 7-isotope network described in \citet{iso7}.  We integrate the
network using the VODE integration package~\citep{vode}, and our
implementation is provided in the StarKiller Microphysics
source~\citep{starkiller}.
We note that we do not explicitly model viscosity.  The reactions will
provide the smallscale cutoff to the instabilities and turbulence at
the flame front.  We also do not include species
diffusion---astrophysical flames tend to have large Lewis numbers, so
this is not expected to be important~\citep{timmeswoosley:1992}.
Finally, we use the thermal conductivities described in
\citet{Timmes00}.

@@

{\color{red} We should decide on a notation for 2D and laminar flame speeds.}

@@

We measure the flame width as:
\begin{equation}
\lambda_f \equiv \frac{\Delta T}{\max\{|\nabla T|\}}
\end{equation}
Experience with modeling resolved flames suggests that we need a
spatial resolution, $\Delta x$ of $\lambda_f/\Delta x \sim 5$~\citep{SNld}.  These
conditions represent the bottom of the He layer.  As the density
decreases with altitude, the flame thickness increases and the flame
speed decreases, so we will easily resolve the flame structure throughout
the rest of the atmosphere. Propagation rates of the 1D laminar flames are given
in Table \ref{table:flame_speeds_1d}.

\begin{deluxetable}{lc}
	\tablecaption{\label{table:flame_speeds_1d} Laminar flame speeds.}
	\tablehead{\colhead{run} & \colhead{$s_L$ (km s$^{-1}$)}}
	\startdata
	10/10 boost & $1.06 \pm 0.01$ \\
	5/5 boost & $0.56 \pm 0.01$ \\
	\enddata
	%
\end{deluxetable}

@@

Table~\ref{table:sim_names} summarizes the simulations that we ran. \MarginPar{This table has different model names from the flame speed table}
The majority of them used a reaction rate boosting of $10$ and a
conductivity boosting of $10$, which should increase the flame speed
by a factor of $10$.  Most simulations used a resolution of
$10~\mbox{cm}$ and a domain width of a little more than one kilometer.  
We use an artifically high rotation rate of
$2000~\mbox{Hz}$, which gives a Rossby length of
\begin{equation}
L_R = \frac{\sqrt{g H_0}}{\Omega} \sim 3\times 10^4~\mathrm{cm}
\end{equation}
using a scale height $H_0 = 10^3~\mathrm{cm}$.  This is about one
quarter of the domain width.  The run at $20~\mbox{cm}$ resolution used
one fewer level of refinement.  The slower rotating case using a
slightly wider domain to accommodate the expected larger Rossby
length.  We note that the entire simulation framework for these
calculations is freely available in the \castro\ github
repository\footnote{\url{https://github.com/AMReX-Astro/Castro}, using
  the {\tt flame\_wave} setup.}.  In the discussions below, we'll use
the simulation names defined in the table to refer to specific runs
and we'll use the 10/10 run as the reference calculation.

@@

Figure~\ref{fig:10_10_overview} shows the temperature, energy
generation rate, and $w$ component of the velocity (the out-of-plane
velocity induced by the Coriolis force).  This latter field
illustrates the hurricane effect set up by the laterally spreading
burning front.  In the energy generation rate plot, we see that the
burning is mostly concentrated toward the bottom of the layer, as
expected since the density is greatest there.  We see that the peak of
the burning has moved off of the symmetry axis, demonstrating that the
burning front is propagating to the right.  In the temperature plot,
we see the effect of our refinement criteria focusing only on the part of
the atmosphere where we are most dense, with an artificial change in the
temperature at the refinement boundary due in part to the construction
of the initial model using the fine grid resolution for HSE. \MarginPar{Remind the reader what this
  was/where they should be looking in the plot to see this temperature
  drop} As we will see in the lower resolution case, we do not believe
that this affects the results.

@@

We also considered the effect of the network size.
Figure~\ref{fig:network} shows a comparison of the 10/10 boosting run
with the standard 13-isotope ``aprox13'' network and the reduced\MarginPar{We use {\tt tt} instead of quotes for the networks in the numerical approach section.}
7-isotope ``iso7'' network.  We see that the flame in the aprox13 case
is slightly more advanced and has a higher $\bar{A}$ than the iso7
case.  We see in the next section that these two networks give largely
the same flame speed.  The reduced network size saves a lot of memory,
which will be useful when we transition to 3-d simulations.

@@

As we see in Figure~\ref{fig:flame_speed}, all of the various
approximations behave as we would expect.  The 1000~Hz run goes 2
times faster than the 2000~Hz run, as expected by the reduced Rossby
radius \citep{cavecchi:2013}).  The 5/5 boost run is half the speed of
the 10/10 boost run, also as expected, consistent with the 1D laminar
flames and behaving as predicted by Eq.~\ref{eq:flame_scaling}.  This
confirms that the role of the Coriolis force is really to boost the
laminar flame by the geometrical/hydrodynamical effect expected ($v_f
= v_l * L_R / H$), similarly to what the 1000 Hz simulation
said.

@@

Baroclinicity is calculated as
\begin{equation}
    \boldsymbol{\psi} = \frac{1}{\rho^2} \nablab p \times \nablab \rho,
\end{equation}
and shows the misalignment of the local density and pressure gradients
\citep{Malone2014a}. As we are considering a 2d \MarginPar{We've used
  at least three different ways of writing 2-d/2D/2d so far. These
  should be changed so we used one style throughout.}system, the
component of the baroclinicity we consider here (out of the plane)
reflects the misalignment of the fields in the plane of the
simulation. Figure~\ref{fig:baroclinicity} shows that the
baroclinicity peaks along the flame front. This baroclinicity
generates vorticity, which in turn entrains material along the surface
of the flame (see also \citet{cavecchi:2013}). In 3D, his same vorticity
should perturb the flame front and further increase the flame speed.
\citep{Cavecchi2019}.

@@

To test our implementation of diffusion, we created a unit test in
\castro\ that just uses the diffusion solver.  This test uses the
standard Gaussian initial conditions---the diffusion of a Gaussian
remains Gaussian, with a lower amplitude and wider width.  This is run
with a constant diffusion coefficient. The point of this test is to
demonstrate that the implementation done in a predictor-corrector
form is second-order accurate.
